---
title: "How has the NBA changed over the years"
author: "Cesar Macias"
date: "2024-04-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

## Our Questions
Main Questions:
How has NBA Offenses changed over the years? 
How different was it 20, 10, and 5 years ago compared to now?

Other Questions:
Are basketball games in the NBA played at a faster pace?
How are the games being played faster?
Are teams shooting more three pointers?
If a team is great offensively, is the best player on that team?
How has the NBA Changed for an individual player?
Are there more great players now? 

## Get Ready
```{r}
library(dplyr)
library(tidyr)
library(tidyverse)
library(ggplot2)

```

## Unfiltered data
```{r}
team_totals <- read.csv("Team Totals.csv")

team_per_game <- read.csv("Team Stats Per Game.csv")
```

## 2004 Top 10 Scoring Teams
```{r}
team_per_game %>%
  filter(season == "2004") %>%
  arrange(desc(pts_per_game)) %>%
  head(10) %>%
  select(season, team, abbreviation, pts_per_game)

```

## 2014 Top 10 Scoring Teams
```{r}
team_per_game %>%
  filter(season == "2014") %>%
  arrange(desc(pts_per_game)) %>%
  head(10) %>%
  select(season, team, abbreviation, pts_per_game)
```

## 2019 Top 10 Scoring Teams
```{r}
team_per_game %>%
  filter(season == "2019") %>%
  arrange(desc(pts_per_game)) %>%
  head(10) %>%
  select(season, team, abbreviation, pts_per_game)
```

## 2024 Top 10 Scoring Teams
```{r}
team_per_game %>%
  filter(season == "2024") %>%
  arrange(desc(pts_per_game)) %>%
  head(10) %>%
  select(season, team, abbreviation, pts_per_game)
```

```{r}

```

